---
title: "Capstone — Rural Hospital Access"
subtitle: "Data visualization for scientists & public-health professionals — carrying one project, rural hospital closures, from a clean table to a presentation-ready figure, in the tidyverse"
author: "Introduction to R"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    df-print: paged
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## What this capstone is about

This notebook collects **both parts** of the course capstone in one
place. It carries a single project — rural hospital closures, and the
access that remains — from a clean table to a presentation-ready figure.

- **Part 1** applies the distribution and category-comparison skills:
  *which states and payment models bore the brunt, and were the
  hospitals that closed big or small?*
- **Part 2** applies the relationship, time-series, faceting, and
  polishing skills: *did closures accelerate, do the hardest-hit states
  carry the heaviest disease burden, and how does access hold up where
  hospitals remain?*

Each part leads with a driving question and closes with an insight plus
one surprise. Each task has a prompt, an empty cell to try it, and a
collapsed **Solution**. Run the setup cell first; everything else
follows from it.

::: callout-tip
## The translation from pandas/seaborn to the tidyverse

If you came to this from the Python version, the moves map across almost
one-to-one:

- **`pandas` → `dplyr`/`readr`.** `read_csv()`, `mutate()`, `filter()`,
  `group_by() |> summarize()`, `count()` do the work of `pd.read_csv`,
  column assignment, boolean masks, and `groupby`.
- **`seaborn`/`matplotlib` → `ggplot2`.** Every chart is the same
  grammar: `ggplot(data, aes(...))` plus a `geom_*`. A seaborn
  *figure-level* call like `relplot(col=...)` becomes
  `facet_wrap(~ ...)`; stacking two plots that share an axis is
  `patchwork`.

So the questions and the pictures are identical; only the dialect
changes.
:::

```{r}
suppressWarnings(suppressMessages({
  library(tidyverse)
}))

# One-time installs for two helpers used in the Part 2 solutions:
#   install.packages("patchwork")   # stack the histogram over the boxplot (Task 2)
#   install.packages("slider")      # trailing rolling mean (Part 2, Task 4)

BASE_URL <- "https://raw.githubusercontent.com/jimcody2014/2026-python-data/refs/heads/main/"
```

------------------------------------------------------------------------

## Capstone · Part 1 — Count and compare the closures

This is the first half of the capstone. It applies the **distribution
and category-comparison** skills to the project that runs through both
courses: **rural hospital closures**. You cleaned and analyzed this data
earlier; here you begin to *show* it.

**Driving question.** Which states and payment models bore the brunt of
rural hospital closures — and were the hospitals that closed big or
small?

We start from the cleaned data. The cell below reproduces, in one step,
the tidy-up from the data-wrangling capstone — dropping the saved index,
fixing the mistyped closure year, standardizing `closure_type`, filling
the few missing bed counts, removing bed-count outliers with the 1.5 ×
IQR rule, and dropping rows with no state — so we begin analysis-ready
and spend our time on the charts.

```{r}
closures <- read_csv(paste0(BASE_URL, "rural_hospital_closures.csv"),
                     show_col_types = FALSE) |>
  select(-1) |>                                       # drop the saved row-index ("Unnamed: 0")
  mutate(
    closure_year = if_else(closure_year == 1019, 2019, closure_year),
    closure_type = str_to_sentence(str_trim(closure_type)),
    closure_type = case_when(
      closure_type == "Complet"  ~ "Complete",        # fix the two typos
      closure_type == "Convertd" ~ "Converted",
      TRUE                       ~ closure_type
    ),
    beds = if_else(is.na(beds), median(beds, na.rm = TRUE), beds)  # fill the few missing beds
  )

# remove bed-count outliers with the 1.5 × IQR rule, then drop stateless rows
qs  <- quantile(closures$beds, c(0.25, 0.75))
iqr <- qs[2] - qs[1]
closures <- closures |>
  filter(between(beds, qs[1] - 1.5 * iqr, qs[2] + 1.5 * iqr)) |>
  drop_na(state)

dim(closures)
```

::: callout-note
## R mechanic: `select(-1)` drops the unnamed index column

When a CSV was saved with its row index, `readr` reads that first column
back in with an auto-generated name like `...1`. `select(-1)` removes it
*by position*, so you don't have to know the name. This is the R
counterpart to pandas' `drop(columns="Unnamed: 0")`.
:::

------------------------------------------------------------------------

### Task 1 — Which states lost the most?

Make a horizontal count plot of the **top 10 states** by number of
closures, ordered most-to-least, fully labeled.

```{r}

```

------------------------------------------------------------------------

### Task 2 — Big hospitals or small?

Show the distribution of `beds` for the hospitals that closed: a
histogram on top, a boxplot beneath, sharing the x-axis. In a comment,
give the median bed count.

```{r}

# Small: median is 25 beds, and the whole distribution sits under about 55.
```

------------------------------------------------------------------------

### Task 3 — Do some payment models lose bigger hospitals?

Plot the **mean `beds` by `payment_type`**, ordered by height, with an
honest zero baseline. (The `payment_types.csv` file in the repo
documents what each code means.)

```{r}

# RRC hospitals that closed are the largest on average (~35 beds); REH the smallest (~19).
```

------------------------------------------------------------------------

### One more look — does closure *type* relate to size?

Before the wrap, one comparison worth drawing honestly: do hospitals
that close *completely* differ in size from those that *convert* to
another kind of facility?

```{r}
closures |>
  group_by(closure_type) |>
  summarize(mean_beds = mean(beds), .groups = "drop") |>
  ggplot(aes(closure_type, mean_beds)) +
  geom_col(fill = "steelblue") +
  labs(title = "Mean beds by closure type (honest baseline)",
       x = "Closure type", y = "Mean beds") +
  theme_minimal()
```

### The insight — and one surprise

**Insight.** Rural closures are not spread evenly. A handful of southern
states — Texas, Oklahoma, Tennessee, Alabama — carry a large share, and
the hospitals that close are overwhelmingly small, with a median of 25
beds. Geography and size, not payment model, tell most of the story.

**One surprise.** Whether a hospital *closed completely* or *converted*
to another kind of facility has almost nothing to do with its size: the
mean bed counts are 29.8 and 28.7 — about one bed apart. On the zero
baseline above, the two bars look identical, which is the honest
reading. It would take a truncated y-axis to inflate that one-bed gap
into a finding — a standing reminder to distrust any version of this
chart whose y-axis does not start at zero.

------------------------------------------------------------------------

## Capstone · Part 2 — Trend, relate, and present the closures

One last return to **rural hospital closures**. In Part 1 you counted
and compared them; now you bring the rest of the visualization toolkit —
trends over time (Lesson 16a), a relationship (Lesson 14), small
multiples (Lesson 15), and presentation polish (Lesson 15) — and finish
with a figure ready for a slide.

**Driving question.** Did rural closures accelerate over time, do the
hardest-hit states carry the heaviest disease burden, and how does
access hold up where hospitals remain?

We start, as before, from the cleaned closures data.

```{r}
closures <- read_csv(paste0(BASE_URL, "rural_hospital_closures.csv"),
                     show_col_types = FALSE) |>
  select(-1) |>
  mutate(
    closure_year = if_else(closure_year == 1019, 2019, closure_year),
    closure_type = str_to_sentence(str_trim(closure_type)),
    closure_type = case_when(
      closure_type == "Complet"  ~ "Complete",
      closure_type == "Convertd" ~ "Converted",
      TRUE                       ~ closure_type
    ),
    beds = if_else(is.na(beds), median(beds, na.rm = TRUE), beds)
  )

qs  <- quantile(closures$beds, c(0.25, 0.75))
iqr <- qs[2] - qs[1]
closures <- closures |>
  filter(between(beds, qs[1] - 1.5 * iqr, qs[2] + 1.5 * iqr)) |>
  drop_na(state)

dim(closures)
```

------------------------------------------------------------------------

### Task 1 — The trend: did closures accelerate?

Count closures per `closure_year` and plot them as a line, fully
labeled. In a comment, say whether closures sped up over the period.

```{r}

# Yes: closures drift upward from a couple a year in the mid-2000s to high single
# digits in the 2010s-2020s, with a sharp spike around 2019.
```

------------------------------------------------------------------------

### Task 2 — The relationship: are the hardest-hit states the sickest?

To relate closures to health, we enrich the closures with county
context. The cell below builds a state-level table — closures per state,
and each state's population-weighted diabetes prevalence from the
ACS/PLACES county data (a `join` and a `group_by`, both from the
data-wrangling course) — so you can focus on the chart.

```{r}
# The two files store the county FIPS code differently, so normalize both keys to
# a zero-padded 5-character string BEFORE joining (see the callout below).

```

::: callout-important
## R mechanic: FIPS codes, leading zeros, and join-key types

This join fails if you read the files as-is. `places$CountyFIPS` stores
county codes with their leading zeros (`"02013"`, `"01001"`), so `readr`
reads it as **character**. But `acs$CountyId` was saved *without* the
leading zero (`1001`), so `readr` reads it as a **number**.
`inner_join()` then errors with *"incompatible types (double vs
character)"* — and even coerced, the values wouldn't line up (`1001` ≠
`01001`). The fix is to force *both* keys to the same zero-padded
5-character string with `str_pad(as.character(x), 5, pad = "0")` before
joining. Any code that keys on ZIP or FIPS codes needs this habit: keep
them as padded strings, never as numbers.
:::

::: callout-note
## R mechanic: `weighted.mean()` is the population weighting

`weighted.mean(hdiabetes, TotalPop)` averages county diabetes prevalence
with each county's population as the weight — big counties count more —
which is the R equivalent of `np.average(..., weights=...)`. The two
`inner_join()`s replace pandas' `merge` and index-aligned `concat`: keep
only counties (then states) present in both tables.
:::

Now plot it. Draw a scatter with a linear trend line of **`diabetes`
(x)** against **`closures` (y)**, fully labeled. In a comment, say
whether states with a heavier diabetes burden tend to have lost more
hospitals.

```{r}

# Positive (about 0.44): states with higher diabetes prevalence tend to have lost
# more rural hospitals -- the places losing access are also the places that most
# need care.
```

------------------------------------------------------------------------

### Task 3 — Small multiples of remaining access

Where hospitals remain, how busy are the clinics? Load the daily
`clinic_visits` and draw the **weekly mean visits** as small multiples —
one panel per `site`.

```{r}

```

```{r}

```

------------------------------------------------------------------------

### Task 4 — The finale: one presentation-ready figure

Bring it home. For the **Hilltop** site, plot daily visits with a
**7-day rolling mean** on top, **annotate the five-day March outage**
with a shaded span and label, set a presentation theme, and save the
figure as a PNG at 150 dpi. This is the chart you would put on a slide.

```{r}

```

------------------------------------------------------------------------

### The insight — and one surprise

**Insight.** Rural closures are not a finished story from the past: they
*accelerated*, climbing from a couple a year to high single digits with
a spike around 2019. And they are not random — they concentrate in the
states carrying the heaviest diabetes burden, so the communities losing
hospitals are the ones that most need chronic-disease care.

**One surprise.** It is diabetes prevalence, more than income, that
tracks closures: the state-level correlation with the diabetes burden
(about 0.44) runs stronger than the one with poverty (about 0.30). The
clearest signal for *where* rural access is disappearing is not simply
how poor a state is, but how sick it is.

**That closes the course.** You started with the grammar of graphics,
worked through distributions, categories, relationships, and time
series, learned to facet and polish — and carried one real project,
rural hospital access, from a clean table all the way to a figure ready
for an audience.
